# ggmosaic

Qiang Zhao Mike Yao-Yi Wang

```{r, include=FALSE}
 # keep this chunk in your .Rmd file
 knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 10)
```
```{r}
library(rmarkdown)
library(rticles)
```

## Overview

This cheat sheet is inspired by the **Chapter 15 Chart: Mosaic** of the edav.info. Instead of using the mosaic function from the package **vcd** to plot the mosaic plot, this cheat sheet shows how to achieve the same output through using **ggmosaic**. 

## Introduction

* Mosaic plot is only for categorical data
* Variables to put in `geom_mosaic`:
  + **weight**: Count/Freq column
  + **x**: product(Y, X2, X1)
  + **fill**: dependent variable Y
  + **conds**: conditional variable


## Order of splits

The mosaic plot follows the hierarchical structure, thus the order of adding variables is very important. Below we will show a step by step splitting by adding one variable at the time. Before going through the example, one must install and call the package **ggplot2** and **ggmosaic**.

```{r cars}
library(ggplot2)
library(ggmosaic)

df_bin=data.frame(Age=c('old','old','old','old','young','young','young','young'),
                    Favorite=c(rep('bubble gum',2),rep('coffee',2),rep('bubble gum',2),rep('coffee',2)),
                    Music=c(rep(c('classical','rock'),4)),
                    Freq=c(1,1,3,1,2,5,1,0))
df_unbin = data.frame(Age =c(rep("old",6), rep("young", 8)), 
                      Favorite = c(rep("bubble gum", 2),rep("coffee", 4), rep("bubble gum", 7), "coffee"),
                      Music = c("classical", "rock", rep("classical", 3), "rock", rep("classical", 2), rep("rock", 5), "classical"))
```

## Splitting on One Variable(binned data)

```{r}
df_bin
```


First, we will show the ggmosaic only split on `Age`:

Important: The `ggmosaic` can take binned data by assigning the weight = `Freq` column of the dataset at its aesthetics, it is not like `vcd::mosaic()`, which can only take binned data with count column name as Freq.
```{r}

ggplot(data = df_bin)+
  geom_mosaic(aes(x = product(Age), fill = Age, weight = Freq))+
  labs(x= "Age", title = "Spliting on Age(binned data)")+
  theme(plot.title = element_text(hjust = 0.5))
  
```


## Splitting on One Variable(unbinned data)

However, for unbinned data, we could just ignore the `weight` and let it set to default.

The unbinned data:
```{r}
df_unbin
```


```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Age), fill = Age))+
  labs(x= "Age", title = "Spliting on Age(unbinned data)")+
  theme(plot.title = element_text(hjust = 0.5))
  
```

Note: We will use unbinned data for the rest of example

## Splitting on Two Variables

Split on `Age`, then `Music`:
```{r}
ggplot(data = df_unbin)+ 
  geom_mosaic(aes(x = product(Music, Age), fill = Music))+
  labs(x = "Age", y = "Music", title = "Spliting on Age, then Music")+
  theme(plot.title = element_text(hjust = 0.5))
```

Split on `Music`, then `Age`:
```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Age, Music), fill = Age))+
  labs(x= "Music", y = "Age",  title = "Spliting on Music, then Age")+
  theme(plot.title = element_text(hjust = 0.5))
```

For plotting mosaic plot on Y ~ X, we want to set `x = product(Y, X)` in `aes` as we always want to split the dependent variable last. We also need to set `fill = Y` as we want to color base on dependent variable.

## Splitting on Three Variables

Split on `Age`, then `Music`, then `Favorite`:

```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Favorite, Music, Age), fill = Favorite))+
  labs(x = "Favorite:Age", y = "Music", title = "Split on Age, then Music, then Favorite")+
   theme(plot.title = element_text(hjust = 0.5))
```

Note that in the above example, by default the order of split and their directions as follows:

1.  `Age` -- vertical split

2.  `Music` -- horizontal split

3.  `Favorite` -- vertical split

## Adjusting the Direction of Splits

The directions can be adjusted as we want. For example, we want to create a doubledecker plot for the above example following below criteria:

Splitting order: 

1.  `Age` -- vertical split ("hspine")

2.  `Music` -- vertical split ("hspine")

3.  `Favorite` (dependent variable)-- horizontal split ("vspine")



```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Favorite, Music, Age), fill = Favorite), 
              divider = c("vspine", "hspine", "hspine"))+
  labs(x = "Music:Age", y = "Favorite", 
       title = "Doubledecker Plot - Split on Age, then Music, then Favorite")+
  theme(plot.title = element_text(hjust = 0.5))

```

Note that the divider vector is in order of which the variables appear in the `product(Favorite, Music, Age)`, however the order of splits is `Age`, `Music`, then `Favorite`. Also note that in the divider vector, "vspine" = horizontal split and "hspine" = vertical split.

## Alternative approach: Conditional
We can also use conditional property to achieve the same result as the above. In this case, `geom_mosaic(aes(x = product(last_split), fill = last_split, conds = product(second_split, first_split))`.

```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Favorite), fill = Favorite, conds = product(Music, Age)), 
              divider = c("vspine", "hspine", "hspine"))+
  labs(x = "Music:Age", y = "Favorite", 
       title = "Doubledecker Plot - (Favorite | Music, Age)")+
  theme(plot.title = element_text(hjust = 0.5))

```


## Alternative approach: Facetting
We can also achieve similar result through facetting.

```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Favorite, Music), fill = Favorite))+
  facet_grid(. ~Age)+
  labs(x="Music", y = "favorite", title = "Favorite ~ Music and facet on Age")+
  theme(plot.title = element_text(hjust = 0.5))

```

## Comparison with vcd::mosaic

There are often confusions between **ggmosaic:geom_mosaic** and **vcd:mosaic** as the syntax for splitting order and splitting direction are quite different for the two. The **vcd:mosaic** follows the order of `mosaic(last_split ~ first_split + second_split)` and the `direction` vector in the order of splits is `(first_split, second_split, third_split)` with "v" being the vertical split and "h" being the horizontal split. However, **ggmosaic:geom_mosaic** follow the different pattern, the order of split is `product(last_split, second_split, first_split)` and the `divider (similar to direction in vcd:mosaic)` in the order of split is `divider = c(last_split, second_split, first_split)` with "vspine" being the horizontal split and "hspine" being the vertical split.

## Chinese Translation: ‘ggmosaic’（马赛克图）

## 引言

这个文档是参考edav.info中第十五章节：马赛克图（Mosaic），引用其中的数据和例子。相较于edav.info中vcd包里的mosiac函数，我们准备使用ggmosiac来画马赛克图。


## 简介
* 马赛克图针对的是分类变量
* geom_maisc中的变量
  + **weight** : 数据中的计数栏
  + **x** : product（因变量，自变量2，自变量1）
  + **fill** : 数据中的自变量栏
  + **conds** : 条件变量

## 分割的顺序
马赛克图遵循等级分层结构，因此往product里面加变量的顺序极其重要。下面，我们会一步一步的展示如何正确的加入变量。我们首先要安装并且引用ggplot2和ggmosaic包。
```{r}
library(ggplot2)
library(ggmosaic)
```

## 根据一个变量分割（分箱数据）：
```{r}
df_bin
```
首先，我们根据年龄（Age）分割：
注意：ggmosaic可以通过weight来处理分箱数据，我们令weight等于数据中的计数栏（Freq）即可。vcd::mosaic也可以做到同样效果，但是计数栏的名称一定要为Freq。
```{r}

ggplot(data = df_bin)+
  geom_mosaic(aes(x = product(Age), fill = Age, weight = Freq))+
  labs(x= "Age", title = "Spliting on Age(binned data)")+
  theme(plot.title = element_text(hjust = 0.5))
  
```

## 根据一个变量分割(非分箱数据):
对于非分箱数据，我们应该忽略weight。下面是非分箱数据的读数:
```{r}
df_unbin
```

```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Age), fill = Age))+
  labs(x= "Age", title = "Spliting on Age(unbinned data)")+
  theme(plot.title = element_text(hjust = 0.5))
  
```
注意：我们接下来的例子都是使用非分箱的数据

## 根据两个变量分割
我们首先根据年龄(Age)分割，然后再根据音乐种类(Music)分割:
```{r}
ggplot(data = df_unbin)+ 
  geom_mosaic(aes(x = product(Music, Age), fill = Music))+
  labs(x = "Age", y = "Music", title = "Spliting on Age, then Music")+
  theme(plot.title = element_text(hjust = 0.5))
```
下面这个例子是先分割音乐种类(Music)，再分割年龄(Age):

```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Age, Music), fill = Age))+
  labs(x= "Music", y = "Age",  title = "Spliting on Music, then Age")+
  theme(plot.title = element_text(hjust = 0.5))
```

如果我们想画因变量Y关于自变量X的马赛克图，我们应该设aes中的x=prod(Y,X)。我们要保证因变量是最后一个被划分的。而且, 我们也要使得fill=Y，因为我们注重的是因变量Y的分布。

## 根据三个变量分割
我们首先划分年龄(Age),然后划分音乐种类(Music)，最后划分喜好(Favorite)。
```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Favorite, Music, Age), fill = Favorite))+
  labs(x = "Favorite:Age", y = "Music", title = "Split on Age, then Music, then Favorite")+
   theme(plot.title = element_text(hjust = 0.5))
```
注意：在上面的例子当中，系统默认的切割方向以及顺序如下:

1. 年龄(Age): 垂直切割
2. 音乐种类(Music): 横向切割
3. 喜好(Favorite): 垂直切割

## 调整切割的方向
我们可以随意改变任意变量的切割方向。比如，我们打算用上面的例子绘画一个双层结构图(DoubleDecker Plot)。
切割顺序为：

1. 年龄(Age): 垂直切割 ('hspine')
2. 音乐种类(Music): 垂直切割 ('hespine')
3. 喜好(Favorite): 横向切割 ('vspine')

```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Favorite, Music, Age), fill = Favorite), 
              divider = c("vspine", "hspine", "hspine"))+
  labs(x = "Music:Age", y = "Favorite", title = "Doubledecker Plot - Split on Age, then Music, then Favorite")+
  theme(plot.title = element_text(hjust = 0.5))

```

注意：divider中的切割方向分别对应product(Favorite,Music,Age)。但是实际上的切割顺序还是年龄(Age)，音乐种类(Music)，喜好(Favorite)。divider中的'vspine'表示横向切割；'hspine'表示垂直切割。

## 另外一种方法:条件变量(Conditional)
我们可以使用条件(conds)的属性来达到与上面例子相同的效果。
geom_mosaic(aes(x = product(最后一次切割), fill = 最后一次切割, conds = product(第二次切割, 第一次切割)).
```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Favorite), fill = Favorite, conds = product(Music, Age)), 
              divider = c("vspine", "hspine", "hspine"))+
  labs(x = "Music:Age", y = "Favorite", 
       title = "Doubledecker Plot - (Favorite | Music, Age)")+
  theme(plot.title = element_text(hjust = 0.5))
```

## 另外一种方法:块化(Facet)
```{r}
ggplot(data = df_unbin)+
  geom_mosaic(aes(x = product(Favorite, Music), fill = Favorite))+
  facet_grid(. ~Age)+
  labs(x="Music", y = "favorite", title = "Favorite ~ Music and facet on Age")+
  theme(plot.title = element_text(hjust = 0.5))

```

## 'ggmosaic' vs vcd::'mosaic'
我们很容易搞混'ggmosaic'和‘mosaic’的切割顺序和切割方向。
对于‘mosaic’来说，切割顺序服从公式'mosaic(最后一个切割~第一个切割+第二个切割)'，而且切割方向的变量也是对应着direction=(第一个切割的变量，第二个切割的变量，最后一个切割的变量)。其中'v'表示的是垂直切割,'h'表示的是横向切割。
但是这和‘ggmosaic’有着很大的不同。在‘ggmosaic’中，切割应服从'product(最后一个切割，第二个切割，第一个切割)'的一个倒叙顺序。而且切割方向的变量也对应着divider=(最后一个切割的变量，第二个切割的变量，第一个切割的变量)。其中'vspine'表示横向切割，‘hspine’表示垂直切割。


