#Running Python code in R Markdown

Karan Rao

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=3.5, fig.height=2)
```

Running Python code in R Markdown can be useful when Python has certain packages or functions R doesn't, or if you're more comfortable performing certain procedures in Python, such as cleaning data or working with strings. Python also has the edge for machine learning packages. One possible workflow could be using Python for cleaning, R for exploratory analysis and plotting, and Python again for machine learning.\
\
Here, we need the `reticulate` package and a Python installation which we can specify by location. We can specify versions of Python with the names of their environments in `use_virtualenv()` and `use_condaenv()`. Instead of the usual `{r}`, we use `{python}` at the start of a Python chunk.

```{r}
library(reticulate)
use_python("/home/karan/anaconda3/bin/python3.5")
```

Objects can be passed easily between R and Python code chunks. To access a Python object in R, prefix it with `py$`. To access an R object in Python, prefix it with `r.` 

```{r}
#R
string_list = c("R", "and", "Python")
```

When an R vector is accessed from a Python code chunk, it automatically becomes a list. The reverse is also true.

```{python}
#Python
sentence = " ".join(r.string_list)
```

```{r}
#R
print(py$sentence)
print(class(py$sentence))
```

Other datatype transformations that occur when objects are passed between R and Python:\
\

<table>
<thead>
<tr>
<th>R</th>
<th>Python</th>
<th>Examples</th>
</tr>
</thead><tbody>
<tr>
<td>Single-element vector&nbsp;&nbsp;</td>
<td>Scalar</td>
<td><code>1</code>, <code>1L</code>, <code>TRUE</code>, <code>"foo"</code></td>
</tr>
<tr>
<td>Multi-element vector</td>
<td>List</td>
<td><code>c(1.0, 2.0, 3.0)</code>, <code>c(1L, 2L, 3L)</code></td>
</tr>
<tr>
<td>List of multiple types</td>
<td>Tuple</td>
<td><code>list(1L, TRUE, "foo")</code></td>
</tr>
<tr>
<td>Named list</td>
<td>Dict</td>
<td><code>list(a = 1L, b = 2.0)</code>, <code>dict(x = x_data)</code></td>
</tr>
<tr>
<td>Matrix/Array</td>
<td>NumPy ndarray</td>
<td><code>matrix(c(1,2,3,4), nrow = 2, ncol = 2)</code></td>
</tr>
<tr>
<td>Data Frame</td>
<td>Pandas DataFrame&nbsp;&nbsp;</td>
<td><code>data.frame(x = c(1,2,3), y = c("a", "b", "c"))</code></td>
</tr>
<tr>
<td>Function</td>
<td>Python function</td>
<td><code>function(x) x + 1</code></td>
</tr>
<tr>
<td>Raw</td>
<td>Python bytearray</td>
<td><code>as.raw(c(1:10))</code></td>
</tr>
<tr>
<td>NULL, TRUE, FALSE&nbsp;</td>
<td>None, True, False</td>
<td><code>NULL</code>, <code>TRUE</code>, <code>FALSE</code></td>
</tr>
</tbody>
</table>
\
 <font size="1">Table source: https://cran.r-project.org/web/packages/reticulate/vignettes/calling_python.html</font>
\
\
If all you need is a couple of lines of Python (or you're writing an R script instead of Markdown), an alternative is the `py_run_string()` function, or the mostly equivalent `py_eval()`. `py_eval()` returns the result of evaluating the expression as opposed to the dictionary associated with the code execution. Both belong to the `reticulate` package.

```{r}
#R
py_run_string("x = ['short', 'order']")
print(py$x)
```
You can also run a Python script with `py_run_file('script.py')`. `source_python('script.py')` is equivalent to importing a Python script. It makes all of its public objects (such as functions) available in the R environment.
\
\
Here's an example that passes a dataframe from R to Python and uses `py_eval()`:
```{r}
#R
data("mtcars")
py_eval("r.mtcars.describe()")
```
\
We can also use Python graphing libraries.
```{python}
import pandas as pd
import seaborn as sns
sns.set(rc={'figure.figsize':(3.5,2)})
sns.distplot(r.mtcars.mpg).set_title("Fuel efficiency")
```
\
\
<font size="4">Working with dictionaries</font>
\
\
Accessing the keys and values of a Python dictionary from R can be confusing at first. This is because Python dictionaries become named lists in R.
```{python}
#Python
cc = {"course":"edav", "year":2019}
```

This is equivalent to:

```{r}
#R
cc_r <- list(course = "edav", year = 2019)
print(names(cc_r))
```

So the original Python object's details can be accessed as below:

```{r}
#R
print(names(py$cc))
#R
print(reticulate::dict(py$cc))
#R
print(reticulate::dict(py$cc)$keys())
```
\
<font size="4">Miscellaneous</font>
\
\
Other packages that allow you to run Python code from within R are [rJython](https://r-forge.r-project.org/projects/rjython/), [rPython](https://cran.r-project.org/web/packages/rPython/index.html), [SnakeCharmR](https://github.com/asieira/SnakeCharmR),  and [PythonInR](https://bitbucket.org/Floooo/pythoninr/).
\
\
Running the `repl_python()` function converts the R console to a Python session, and the objects you create are persistent and can be accessed with the usual `py$` later. Type `exit` to return to the R console when you're done. This also belongs to the `reticulate` package.
\
\
<font size="4">Limitations and troubleshooting</font>
\
\
In older versions of RStudio (before 1.2.1114), running Python chunks directly in the IDE to see output is not supported - they can only be run while knitting. To check if your output works, you can use `repl_python()` as a workaround. You can also put your code in `py_run_string()` in an R chunk.
\
\
It's best to use the latest version of RStudio, which clears up issues related to loading the right Python environments and newly installed Python libraries.
\
\
<font size="4">Sources</font>
\
\
[Reticulate on RStudio's GitHub page](https://rstudio.github.io/reticulate/)\
[Reticulate vignettes](https://cran.r-project.org/web/packages/reticulate/vignettes/calling_python.html)\
[InfoWorld - How to run Python in R](https://www.infoworld.com/article/3340120/how-to-run-python-in-r.html)\
[Running Python from R with Reticulate by Len Kiefer](http://lenkiefer.com/2019/01/28/running-python-from-r-with-reticulate/)\
[From ‘R vs Python’ to ‘R and Python’ - Towards Data Science](https://towardsdatascience.com/from-r-vs-python-to-r-and-python-aa25db33ce17)\
[Snakes in a Package](https://www.mango-solutions.com/blog/snakes-in-a-package-combining-python-and-r-with-reticulate)\
[R Markdown: The Definitive Guide by Xie, Allaire & Grolemund](https://bookdown.org/yihui/rmarkdown/language-engines.html#python)\
[Listen Data - Run Python from R](https://www.listendata.com/2018/03/run-python-from-r.html)\
[StackOverflow - calling Python methods from R](https://stackoverflow.com/questions/51108722/how-to-call-python-method-from-r-reticulate)\
[StackOverflow - running Python chunks in RStudio](https://stackoverflow.com/questions/49503195/reticulate-running-python-chunks-in-rmarkdown)\


